Computational methods in landscape ecology

### 2024-07-02 Manuscript.qmd
Rename files
render MS

---
title: "Computational methods in landscape ecology"
keywords: keyword1, keyword2
author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
    attributes:
      - corresponding: boolean
  - name: Jakub Nowosad
    orcid: 0000-0002-1057-3721
    email: nowosad.jakub@gmail.com
    affiliations:
      - name: Adam Mickiewicz University, Poznan, Poland
  - name: Marti Bosch
    orcid: 0000-0001-8735-9144
    email: marti.bosch@epfl.ch
    affiliations:
      - name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
  - name: Martin Jung
    orcid: 0000-0002-7569-1390
    email: jung@iiasa.ac.at
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
  - name: Rafael Schouten
    orcid: 0000-0002-8380-0884
    email: rafael.schouten@sund.ku.dk
    affiliations:
      - name: Globe Institute, University of Copenhagen, Copenhagen, Denmark
format:
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header: 
      text: | 
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
papersize: a4paper
fontsize: 12pt
geometry: margin=25mm
indent: 2m
linestretch: 1.5
bibliography: CLER.bib
csl: springer-basic-author-date.csl
---

@@

{{< include Sections/introduction.qmd >}}
{{< include Sections/simulations.qmd >}}

@@

### References

::: {#refs}
:::

>>>>

---
title: "Computational methods in landscape ecology"
keywords: keyword1, keyword2
author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
    attributes:
      - corresponding: boolean
  - name: Jakub Nowosad
    orcid: 0000-0002-1057-3721
    email: nowosad.jakub@gmail.com
    affiliations:
      - name: Adam Mickiewicz University, Poznan, Poland
  - name: Marti Bosch
    orcid: 0000-0001-8735-9144
    email: marti.bosch@epfl.ch
    affiliations:
      - name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
  - name: Martin Jung
    orcid: 0000-0002-7569-1390
    email: jung@iiasa.ac.at
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
  - name: Rafael Schouten
    orcid: 0000-0002-8380-0884
    email: rafael.schouten@sund.ku.dk
    affiliations:
      - name: Globe Institute, University of Copenhagen, Copenhagen, Denmark
format:
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header: 
      text: | 
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
papersize: a4paper
mainfont: "Garamond"
fontsize: 11pt
geometry: margin=25mm
indent: 2m
linestretch: 1.5
bibliography: CLER.bib
csl: springer-basic-author-date.csl
---

@@

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/05-simulations.qmd >}}

@@

### References {.unnumbered}

\begingroup
\raggedright
::: {#refs}
:::
\endgroup

### 2024-07-02 Manuscript.tex
render MS

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
  a4paperpaper,
]{article}

@@

\subsection{Landscapes generators}\label{landscapes-generators}

Landscape generators are typically used as null hypotheses, baselines,
or scenarios in landscape ecology analyses because they allow to control
certain aspects of the landscape (Li et al. 2004; Wang and Malanson
2008) and are generally simpler as ecological simulations models (Hess
et al. 2020). Landscape generators can be classified into two major
categories, namely pattern-based and process-based approaches (Pe'er et
al. 2013).

Pattern-based landscape generators simulate landscape patterns without
assuming any underlying abiotic or biotic processes (i.e., neutral
landscape model; Gardner et al. 1987). The earliest neutral models are
based on percolation theory and assign LULC types randomly to cells in
the landscape(Gardner et al. 1987) or include hierarchical models that
consider different spatial scales while assigning cell values (O'Neill
et al. 1992). Landscapes characterized by continuous values can be
simulated by fractal models, such as Brownian motion (Palmer 1992).
Borrowing from computer graphics and being able to simulate hierarchical
landscapes, more recent neutral landscape models make use of spectral
synthesis (e.g., Perlin noise; Etherington 2022) or Binary space
partitioning (Etherington et al. 2022). Neutral landscape models are
also able to simulate landscapes dominated by anthropogenic activities,
e.g., based on least-cost networks (Etherington et al. 2024). In order
to ensure realistic landscapes, several approaches exist that use real
landscapes (Inkoom et al. 2017) or are based on target values
optimization (Van Strien et al. 2016; Justeau-Allaire et al. 2022) to
simulate neutral landscapes.

Contrastingly, processes-based landscape generators explicitly include
hypothesized pattern creating processes (Langhammer et al. 2019;
Salecker et al. 2019).

\subsection{IBMs /Coupled lattice}\label{ibms-coupled-lattice}

@@

\subsubsection{References}\label{references}

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
  a4paperpaper,
]{article}

@@

\subsection{Landscapes simulators}\label{landscapes-simulators}

Landscape simulators are typically used as null hypotheses, baselines,
or scenarios because they allow to control certain aspects of the
landscape (Li et al. 2004; Wang and Malanson 2008) and are generally
simpler than ecological simulations models (Hess et al. 2020). Landscape
simulators can be classified into two major categories, namely
pattern-based and process-based approaches {[}Pe'er et al. (2013);
Langhammer2019{]}.

Pattern-based approaches simulate landscapes without assuming any
underlying abiotic or biotic processes (i.e., neutral landscape model;
Gardner et al. 1987). The earliest neutral models are based on
percolation theory and assign LULC types randomly to cells in the
landscape(Gardner et al. 1987) or hierarchical models that consider
different spatial scales while assigning cell values (O'Neill et al.
1992). Landscapes characterized by continuous values can be simulated by
fractal models, such as Brownian motion (Palmer 1992). Borrowing from
computer graphics, more recent neutral landscape models make use of
spectral synthesis (e.g., Perlin noise; Etherington 2022) or binary
space partitioning (Etherington et al. 2022). Neutral landscape models
are also able to simulate landscapes dominated by anthropogenic
activities using based on least-cost networks (Etherington et al. 2024).
In order to ensure realistic neutral landscapes several approaches exist
that use comparisons with real landscapes (Inkoom et al. 2017) or are
based on target values optimization (Van Strien et al. 2016;
Justeau-Allaire et al. 2022).

Contrastingly, processes-based approaches explicitly include abiotic or
biotic pattern-forming processes (i.e., landscape generators; Langhammer
et al. 2019; Salecker et al. 2019). Several earlier landscape generators
are based on cellular automata models and are able to simulate urban
growth (Clarke and Gaydos 1998) or deforestation (Soares-Filho et al.
2002). More recent landscape generators allow to simulate patchy
landscapes based on Gibbs processes (Gaucherel et al. 2006),
deforestation based on road and agricultural fields access (Pe'er et al.
2013), agricultural areas dominated by smallholders (Salecker et al.
2019), or vegetation surround watering points in semi-arid savanna
rangelands (combining pattern- and process-bases approaches; Hess et al.
2020).

\subsection{Individual-based models}\label{individual-based-models}

Individual-based models (or agent-based models) simulate discrete
entities that are described by attributes and behaviors and patterns
emerge from bottom-up interactions of individuals with each other and
their environment (DeAngelis and Grimm 2014). In landscape ecology
individual-based models are increasingly used to model especially
social-ecological systems (e.g, Carauta et al. 2018; Dislich et al.
2018; Gonzalez-Redin et al. 2019), but also disturbances (e.g., Seidl
and Rammer 2017; Spies et al. 2017; Kamimura et al. 2019) or
connectivity (e.g., Day et al. 2020). In order to facilitate future
model developments, recent progress includes adaptable modeling
frameworks (Schumaker and Brookes 2018) or reusable building block
(Berger et al. 2024). Furthermore, hybrid approaches that couple or even
integrated different model types with individual-based models could
benefit from rich development history or facilitate corresponding
strengths (O'Sullivan et al. 2016), e.g., linking an individual-based
and ecosystem model (Polhill et al. 2011).

@@

\subsubsection*{References}\label{references}
\addcontentsline{toc}{subsubsection}{References}

\begingroup
\raggedright

@@

\bibitem[\citeproctext]{ref-Berger2024}
Berger U, Bell A, Barton CM, et al (2024) Towards reusable building
blocks for agent-based modelling and theory development. Environmental
Modelling \& Software 175:106003.
\url{https://doi.org/10.1016/j.envsoft.2024.106003}

@@

\bibitem[\citeproctext]{ref-Carauta2018}
Carauta M, Latynskiy E, Mössinger J, et al (2018) Can preferential
credit programs speed up the adoption of low-carbon agricultural systems
in mato grosso, brazil? Results from bioeconomic microsimulation.
Regional Environmental Change 18:117--128.
\url{https://doi.org/10.1007/s10113-017-1104-x}

@@

\bibitem[\citeproctext]{ref-Clarke1998}
Clarke KC, Gaydos LJ (1998) Loose-coupling a cellular automaton model
and GIS: Long-term urban growth prediction for san francisco and
washington/baltimore. International Journal of Geographical Information
Science 12:699--714. \url{https://doi.org/10.1080/136588198241617}

@@

\bibitem[\citeproctext]{ref-Day2020}
Day CC, Zollner PA, Gilbert JH, McCann NP (2020) Individual-based
modeling highlights the importance of mortality and landscape structure
in measures of functional connectivity. Landscape Ecology 35:2191--2208.
\url{https://doi.org/10.1007/s10980-020-01095-5}

\bibitem[\citeproctext]{ref-DeAngelis2014}
DeAngelis DL, Grimm V (2014) Individual-based models in ecology after
four decades. F1000Prime Reports 6:1--6.
\url{https://doi.org/10.12703/P6-39}

\bibitem[\citeproctext]{ref-Dislich2018}
Dislich C, Hettig E, Salecker J, et al (2018) Land-use change in oil
palm dominated tropical landscapes - an agent-based model to explore
ecological and socio-economic trade-offs. PLoS ONE 13:e0190506.
\url{https://doi.org/10.1371/journal.pone.0190506}

@@

\bibitem[\citeproctext]{ref-Gaucherel2006}
Gaucherel C, Fleury D, Auclair D, Dreyfus P (2006) Neutral models for
patchy landscapes. Ecological Modelling 197:159--170.
\url{https://doi.org/10.1016/j.ecolmodel.2006.02.044}

@@

\bibitem[\citeproctext]{ref-Gonzalez-Redin2019}
Gonzalez-Redin J, Gordon IJ, Hill R, et al (2019) Exploring sustainable
land use in forested tropical social-ecological systems: A case-study in
the wet tropics. Journal of Environmental Management 231:940--952.
\url{https://doi.org/10.1016/j.jenvman.2018.10.079}

@@

\bibitem[\citeproctext]{ref-Kamimura2019}
Kamimura K, Gardiner B, Dupont S, Finnigan J (2019) Agent-based
modelling of wind damage processes and patterns in forests. Agricultural
and Forest Meteorology 268:279--288.
\url{https://doi.org/10.1016/j.agrformet.2019.01.020}

@@

\bibitem[\citeproctext]{ref-OSullivan2016}
O'Sullivan D, Evans T, Manson S, et al (2016) Strategic directions for
agent-based modeling: Avoiding the YAAWN syndrome. Journal of Land Use
Science 11:177--187. \url{https://doi.org/10.1080/1747423X.2015.1030463}

@@

\bibitem[\citeproctext]{ref-Polhill2011}
Polhill JG, Gimona A, Aspinall RJ (2011) Agent-based modelling of land
use effects on ecosystem processes and services. Journal of Land Use
Science 6:75--81. \url{https://doi.org/10.1080/1747423X.2011.558603}

@@

\bibitem[\citeproctext]{ref-Schumaker2018}
Schumaker NH, Brookes A (2018) HexSim: A modeling environment for
ecology and conservation. Landscape Ecology 33:197--211.
\url{https://doi.org/10.1007/s10980-017-0605-9}

\bibitem[\citeproctext]{ref-Seidl2017a}
Seidl R, Rammer W (2017) Climate change amplifies the interactions
between wind and bark beetle disturbances in forest landscapes.
Landscape Ecology 32:1485--1498.
\url{https://doi.org/10.1007/s10980-016-0396-4}

@@

\bibitem[\citeproctext]{ref-Soares-Filho2002}
Soares-Filho BS, Coutinho Cerqueira G, Lopes Pennachin C (2002)
Dinamica---a stochastic cellular automata model designed to simulate the
landscape dynamics in an amazonian colonization frontier. Ecological
Modelling 154:217--235.
\url{https://doi.org/10.1016/S0304-3800(02)00059-5}

\bibitem[\citeproctext]{ref-Spies2017}
Spies TA, White E, Ager A, et al (2017) Using an agent-based model to
examine forest management outcomes in a fire-prone landscape in oregon,
USA. Ecology and Society 22:art25.
\url{https://doi.org/10.5751/ES-08841-220125}

@@

\endgroup
